\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage{hyperref}

\title{Lab 1B Written Answers}
\author{}
\date{}

\begin{document}
\maketitle

\section*{Part 2 --- Training Choices (Discussion)}

\subsection*{Learning rate choices}
A learning rate that is too large makes optimization unstable: the loss oscillates and the decision boundary can jump around between epochs. A learning rate that is too small converges very slowly, especially on the spiral dataset where fine curved boundaries are required. In these experiments, values around 0.01 worked well for small/medium models, while wider networks were more stable around 0.005.

\subsection*{Optimizer choice (Adam vs SGD)}
Adam was a good default because it adapted per-parameter step sizes and converged quickly on all three datasets with minimal tuning. Plain SGD can still work, but it usually needs more careful learning-rate scheduling and often more epochs to match Adam performance. Since the lab emphasizes architecture comparison rather than optimizer tuning, Adam keeps the comparison cleaner.

\subsection*{Epoch count and convergence}
Blobs typically converged fastest because the classes are close to linearly separable. Ellipsoids required moderate training because overlap introduces ambiguity near class boundaries. Spirals needed the most epochs because the model must learn highly nonlinear, intertwined boundaries.

\subsection*{Sensitivity to initialization}
Shallow/narrow models showed stronger initialization sensitivity on spirals: some starts converged to mediocre boundaries. Deeper or wider networks were generally more robust but still benefited from stable learning rates. Good initialization mainly reduced time-to-convergence and prevented early saturation.

\subsection*{Depth/width and training speed}
Increasing width often improved fitting capacity quickly but raised compute cost per epoch. Increasing depth improved representational flexibility for curved boundaries, but deeper models could train slower and be slightly harder to tune. In practice, moderate depth with adequate width gave the best trade-off.

\section*{Part 2 --- Reflection Questions}
\begin{enumerate}[label=\textbf{\arabic*.}]
  \item \textbf{Why does adding depth help on the spiral dataset but not on the blobs dataset?}\\
  Blobs are close to linearly separable, so even simple models already capture the boundary well. Spirals require multiple nonlinear transformations to ``untangle'' class manifolds, which depth provides by composing several learned feature mappings. Therefore, depth adds little on easy linear structure but helps significantly on complex geometry.

  \item \textbf{Why might very wide networks succeed even when they have only one or two hidden layers?}\\
  Width gives many parallel neurons that can approximate rich nonlinear functions even with limited depth. A sufficiently wide hidden layer can carve complex decision regions by combining many local features. This can compensate for less hierarchical feature composition.

  \item \textbf{Which dataset was the hardest for the MLP, and what properties of the dataset made it difficult?}\\
  The spiral dataset was hardest. Its classes are intertwined and non-convex, so good performance requires curved, high-complexity decision boundaries and careful optimization. Small or undertrained models tend to underfit this geometry.

  \item \textbf{How did depth and width affect training, in terms of sensitivity to initialization and training time?}\\
  More depth generally increased expressiveness but also made optimization slightly more sensitive and slower per full convergence. More width usually improved fitting and robustness but increased per-epoch computation. Moderate depth plus moderate-to-high width was the most stable configuration.

  \item \textbf{Did deeper networks always perform better? Why or why not?}\\
  No. On simple datasets, extra depth provides little benefit and may only add optimization overhead. Performance depends on matching model capacity to data complexity and training setup; beyond that point, deeper models can plateau or occasionally underperform if not tuned well.
\end{enumerate}

\section*{Part 3 --- Reflection Questions}
\begin{enumerate}[label=\textbf{\arabic*.}]
  \item \textbf{Why does an MLP without nonlinearities collapse into a linear model?}\\
  A composition of linear maps is still a linear map. If every layer is linear, the whole network is equivalent to one matrix multiplication plus bias, so it cannot represent nonlinear class boundaries.

  \item \textbf{What would happen if ReLU backward incorrectly returned gradients for negative inputs?}\\
  Neurons that should be inactive would still receive updates, which changes the effective function and breaks the true derivative. Training could become biased or unstable, and the model may converge to worse solutions because gradients no longer match the forward computation.

  \item \textbf{Why is it useful to wrap the forward/backward logic inside \texttt{torch.autograd.Function} instead of writing manual gradient loops?}\\
  \texttt{torch.autograd.Function} integrates custom ops into PyTorch's computation graph cleanly and efficiently. It keeps tensor operations vectorized, allows automatic chaining with other layers, and avoids error-prone manual per-parameter loop code. This is the standard way to define custom differentiable operations in PyTorch.
\end{enumerate}

\end{document}
